---
title: "Statistical Rethinking (UiO PSY9550):\nOrganisation and overview"
author: "Guido Biele"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
editor: source
---

Seminar based on Richgard McElreath's book [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/)

## Who am I? Who are you?

# Organisation

## Learning goals

- You understand Bayesian analysis as an alternative to NHST
- You think more clearly about causation
- You understand simple and moderately complex regression models
- You learn how to learn on your own

## Prerequisits

- Some R knowledge
- Willing to spend a day a week preparing

## Procedure

1. Read weekly chapter
2. Solve exercises and submit html created from quarto Rmarkdown (in teams)
3. Submit solutions until Wednesday noon
4. (GB summarizes material)
5. Present results in class (one team per question)

## Teams

Solving the exercises together is more fun and you can help each other. It also lessens the burden.
Find teams of 2 people and think about a name. 

_Email me if you are without a team by Monday evening._

## Submitting solutions

[here](https://github.com/gbiele/StatRethink/tree/master/Orga) you find Rmarkdown and quarto templates that you can use to prepare your the answers you submit.

To download such a template, right-click either here [quarto](https://raw.githubusercontent.com/gbiele/StatRethink/master/Orga/qSkeleton.qmd) here [Rmarkdown](https://raw.githubusercontent.com/gbiele/StatRethink/master/Orga/Skeleton.Rmd). As quarto is the more modern format, it is a good idea to use that. TO get started, download the skeleton, open it in Rstudio and click the "Render" (or "Knit" for Rmarkdown) button to create the html document for easy reading. To prepare exercise solutions for submission, copy and rename a skeleton, write in your solutions, and render the document.

**Please do not send me the .qmd or .Rmd document, but the .html document that you created.**

Send the solutions via email to guido-point-biele-at-fhi-point-no with the subject line following this pattern: StatRethink Exercises Chapter [chapter] Team [team name].

## Credit points

- For each chapter teams of two students have to submit answers
  - submission by Monday noon
  - All easy questions for each chapter
  - 2-4 Selected medium hard question per chapter
  - 3 hard questions out of any chapter
- Credit points based on submitted solutions

The practice exercises that should be submitted are at the bottom of this page.

## Procedure for each session

1. Summary of chapter
2. Presentation of solutions
3. Questions


# Using ChatGPT

Do it!

But you will be asked to explain your results.
ChatGPT can be very useful for explaining code, or explaining a concept a bit differently. Example prompts.

> You are an expert in Bayesian statistics and are particularily influenced by Statistician Andrew Gelman and Richard McElreaths book "statistical rethinking. Please explain the concept of a prior predictive check.

> Explain the following equation: Pr(W, L, p) = Pr(W, L|p) Pr(p)

> Show a linear regression formulated in ulam from the R package rethinking.

> explain this code from the solutions to excercises of the book "Statistical Rethinking":

```{r}
#| eval: false
m <- quap(
  alist(
    height ~ dnorm( mu , sigma ) ,
    mu <- a + b*( weight - xbar ) ,
    a ~ dnorm( 178 , 20 ) ,
    b ~ dlnorm( 0 , 1 ) ,
    sigma ~ dunif( 0 , 50 )
  ) , data=d2 )

```

Use chatGPT as tutor, not as a fast way to complete exercises.


# Installing the `rethinking` package

[Here](https://github.com/rmcelreath/rethinking) are the official instructions.

## Install rethinking slim (easy)

This is sufficient for the first 6 or so weeks.

```{r}
#| eval: false
install.packages(c("coda","mvtnorm","devtools","loo","dagitty"))
devtools::install_github("rmcelreath/rethinking@slim")
```

## Install rstan and cmdstanr (harder)

We'll need this from around week 7 on.
- needs administrator rights
- [Instructions for rstan](https://mc-stan.org/users/interfaces/rstan.html) 
- [Instructions for cmdstanr](https://mc-stan.org/cmdstanr/) 

If you do not have admin rights on your computer, it is best to do the installation together with your IT-help.

# Videos

Richard McElreath has lectures on [youtube](https://www.youtube.com/watch?v=FdnMWdICdRs&list=PLDcUM9US4XdPz-KxHM4XHt7uUVGWWVSus). **These cannot replace reading the chapters!**

# Overview

- Bayesian updating and probability distributions (Chapters 2-4)
- Causal inference (Chapters 5,6)
- Information theory and model comparison (Chapter 7)
- Markov Chain Monte Carlo sampling (Chapter 9)
- Regression model and Causal inference (Chapters 8, 10-13)
  - interactions
  - logistic & ordinal regression
  - multilevel/hierarchical regression
    
We skip some chapters (10) and do others only partially (9, 11, 12) 

# Exercises

- Chapter 1 & 2: 2E1-2E4, 2M1-2M3