---
title: "Exercises for chapter x"
author: "team name"
date: "date"
output:
  html_document: 
    mathjax: default
    toc: true
    toc_depth: 2
header-includes: 
    \usepackage{xcolor}
    \usepackage{amsmath}
---

```{css, echo = F}
body{
  font-family: Helvetica;
  font-size: 16pt;
  max-width: 1000px;
  margin: auto;
  margin-left:310px;
}
pre{
  font-size: 14px;
}
/* Headers */
h1{
    font-size: 24pt;
  }
h1,h2{
    font-size: 20pt;
  }
h3,h4,h5,h6{
  font-size: 18pt;
}

#TOC {
  position: fixed;
  left: 0;
  top: 0;
  width: 300px;
  height: 100%;
  overflow:auto;
}
```

```{r setup, include=FALSE, message=FALSE, warning=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE, dpi = 300)

library(rethinking)
library(magrittr)
# add other packages you use here
```

# Introductory tutorial

[Here](https://ourcodingclub.github.io/tutorials/rmarkdown/) you find a introductory tutorial for Rmarkdown.

# Chapter 1 exercise E1

Use header 1 style (`# [your header text]`) to indicate the exercise chapter and number.

## Plotting

If necessary, you can use a header 2 style (`# [your header text]`) to indicate sub-section of a solution. This will typically not be necessary.

Plotting is very easy. Just write the code for the plot into a code block:

```{r}
hist(rnorm(1000, mean = 0, sd = 1),
     main = "Histogram of 1000 random draws\nfrom the standard normal distribution")
```

```{r}
curve(dnorm(x, mean = 0, sd = 1),
     from = -4, to = 4, n = 500,
     main = "Probability density for the standard normal distribution")
```

```{r}
curve(pnorm(x, mean = 0, sd = 1),
     from = -4, to = 4, n = 500,
     main = "Cumulative density for the standard normal distribution")
```

```{r}
curve(qnorm(x, mean = 0, sd = 1),
     from = 0, to = 1, n = 500,
     main = "Quantile function for the standard normal distribution")
```

## Hide standard output

If you estimate models with `ulam` it is a good idea to do estimation and plotting of model results in separate code blocks.

Use a code block like a following to estimate the model. Here `results='hide'` makes that the standard output will not be included in the rendered html document.

```{r warning=FALSE, results='hide'}
n.cores = ifelse(Sys.info()["sysname"] == "Darwin",4,1)

my.fit = ulam(
  alist(
    y ~ dnorm(mu, sigma),
    mu ~ dnorm(0,3),
    sigma ~ dexp(2)
  ),
  data = list(y = rnorm(25)),
  log_lik = TRUE,
  chains = 4)
```

And do additional steps with the resulting fit object in the following code blocks:

```{r}
precis(my.fit)
```

```{r}
divergent(my.fit)
```

## Equations

If you want to show equations, it is sufficient to write for example y = a + b\*x.

If you want show prettier equations, have to use LATEX. For instance `$y = a + b*x$`, where the `$`s indicate start and end of an inline equation, becomes $y = a + b*x$. [Here](https://tilburgsciencehub.com/building-blocks/collaborate-and-share-your-work/write-your-paper/amsmath-latex-cheatsheet/) is a cheat sheet that explains latex commands.

For instance, we can write

-   Greek letter like this `$y = \alpha + \beta*x$`: $y = \alpha + \beta*x$.

-   subscripts by using ``` _``  and superscripts by using ```\^`;`$y_i = \alpha + \beta*x_i + \gamma^2$\`: $y_i = \alpha + \beta*x_i + \gamma^2$.

-   fractions like with `\frac{ ... }{ ... }`: `$P(\theta|data) = \frac{P(data|\theta)P(\theta)}{P(data)}$.` becomes $P(\theta|data) = \frac{P(data|\theta)P(\theta)}{P(data)}$. this is Bayes rule. The probability of a hypothesis or parameter value $\theta$ given the data is equal to the probability of the data given $\theta$ times the prior probability of the $\theta$, divided by the probability of the data.

If we have multi line expressions, we start and end with two `$` and ude `\\` to indicate line breaks:

    $$ 
    y \sim normal(\bar \mu, \sigma)   \\ 
    \bar \mu \sim normal(0,3)  \\ 
    \sigma \sim exponential(1) \\
    $$

becomes

$$ 
y \sim normal(\bar \mu, \sigma)   \\ 
\bar \mu \sim normal(0,3)  \\ 
\sigma \sim exponential(1) \\
$$

# Clarification and or discussion questions

Please put things you find unclear / hard to understand / worthy of a discussion here.
